Supplementary Information

### 2021-05-05 doc/manuscript/ForestGEO_treerings.Rmd
PZ comments / cleanup

5. Midwest Dendro LLC, Naperville, Illinois 60565 USA;

** Current address: Decision and Infrastructure Sciences, Argonne National Laboratory; Lamont, IL.

@@

Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., REFS, @evans_fusing_2017; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).
*Moreover, while tree-ring data has been used for benchmarking dynamic global vegetation models [DGVMs, e.g., klesse_combined_2018], an individual-based analysis approach focusing on multiple drivers including DBH would offer more opportunities for using tree-ring data as benchmarking tools for demographic DGVMs [[REFS?](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/128)].*

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
Over the vast majority of Earth's forested regions, water is the primary climatic factor limiting tree growth, whereas low temperatures are the strongest limitation in some high latitude or high elevation sites [@babst_twentieth_2019]. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., **REFS**, @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
First, moisture and temperature are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, REF] are commonly used to assess the joint influence of moisture and temperature, we hypothesize hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`) such that their effects cannot be fully captured by a single metric such as SPEI.
In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @trouillier_size_2019; @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

## DBH
Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @enquist_global_2002; @niklas_plant_2004; **REF**] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
Despite the fundamental importance of a solid understanding of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
In addition, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

Rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings. 
Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
Yet, even after correcting for ontogenetic growth trends, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019], fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all points throughout the history of a dynamically changing stand.    
Stand dynamics strongly influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, affecting ecosystem level patterns of C allocation. 
Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020]. 
Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past couple centuries) will generally decline, whereas populations sampled from old-growth stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).

>>>>

5. Midwest Dendro LLC, Naperville, Illinois 60565 USA; 
** Current address: Decision and Infrastructure Sciences, Argonne National Laboratory; Lamont, IL.

@@

Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., **REFS**, @evans_fusing_2017; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
Over the vast majority of Earth's forested regions, water is the primary climatic factor limiting tree growth, whereas low temperatures are the strongest limitation in some high latitude or high elevation sites [@babst_twentieth_2019]. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., **REFS**, @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
First, moisture and temperature are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, **REF**] are commonly used to assess the joint influence of moisture and temperature, we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`) such that their effects cannot be fully captured by a single metric.
In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @trouillier_size_2019; @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @enquist_global_2002; @niklas_plant_2004; **REF**] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
Despite the fundamental importance of a solid understanding of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
In addition, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

Beyond the direct effects of climate, rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings. 
Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
Yet, even after correcting for ontogenetic growth trends, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019], fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all points throughout the history of a dynamically changing stand.
Stand dynamics strongly influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, affecting ecosystem level patterns of C allocation. 
Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020]. 
Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past couple centuries) will generally decline, whereas populations sampled from old-growth stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
